Manuscript Figures

### 2021-06-10 src/manuscript_figures.Rmd
update MSI panel with benchmarks

# survival prediction accuracy benchmarking
# find survival-predictive LFs by univariate cox models and use them in RSF
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("survival benchmark for ",tool)
  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    require(survminer)
    M <- LFs[[tool]][[experiment]]
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    # compute pfi stats
    # fit a first model to filter variables 
    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
    # using a range of top variables, compute c index
    dt1 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt1$endpoint <- 'PFI'
    # do the same for overall survival
    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
    dt2 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt2$endpoint <- 'OS'
    return(rbind(dt1, dt2))
  }))
}))
parallel::stopCluster(cl)

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.cluster_stability_metrics.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),        
       plot = Fig, units = 'in', 
       width = 10, height = 10)
Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B). Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) based on the top 10 LFs/PCs derived from each tool. C) Same as in panels A and B), but split by the value of k used in clustering. D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival or Progression Free Survival endpoints."
```

@@

# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
event <- 'OS'
time <- 'OS.time'

@@

M <- LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']]
M <- scale(M) #remove_redundant_variables(M, perc = 99)
y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
M <- M[!is.na(y),]
y <- y[!is.na(y)]

res <- predict_cluster_labels_glmnet(M, y)
p1 <- ggplot(res, aes(x = label, y = auc)) +
  geom_bar(stat = 'identity', aes(fill = label)) + 
  labs(y = 'AUC', x = '') +
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2)+
  theme(legend.title = element_blank())

# also plot a roc curve
rocs <- predict_cluster_labels_glmnet(M, y, returnROC = TRUE)
p2 <- plot_roc_list(rocs)

# explore MSI related LFs
# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 10)[['variable']])

df <- plot_tsne(M[,top_vars], y, returnData = T)
# color tsne by top lf informative for MSI-H
top_lf <- dt[order(padj)][ref_cl == 'MSI-H']$variable[1]
df <- merge(df, data.frame(M[,top_lf,drop =F]), by = 'row.names')
df$cancertype <- surv[match(df$Row.names, bcr_patient_barcode)]$project
df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected

p3.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p3.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')

@@

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p4 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + 
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45)) + 
  labs(x = paste0("top features contributing to ",top_lf))

Fig <- cowplot::plot_grid(plot_grid(p1, p2, ncol = 1, labels = c('A', 'B')),
                          plot_grid(p3.1, p3.2, p4, ncol = 1, labels = c('C', 'D', 'E')),
                          nrow = 1)
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.MSI.status.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 8, height = 8)
Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors. A) AUC values for predicting MSI status using MAUI LFs B) Top features that contribute to MAUI latent factor (‘LF95’) that is the most predictive latent factor in distinguishing samples with high microsatellite instability from the rest. C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor (‘LF95’) which is most predictive of MSI-H samples."
```

>>>>

# survival prediction accuracy benchmarking
# find survival-predictive LFs by fitting an initial RSF and use 
# top factors for a final model to compute Harrel's c-index
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("survival benchmark for ",tool)
  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    require(survminer)
    M <- LFs[[tool]][[experiment]]
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    # compute pfi stats
    # fit a first model to filter variables 
    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
    # using a range of top variables, compute c index
    dt1 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt1$endpoint <- 'PFI'
    # do the same for overall survival
    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
    dt2 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt2$endpoint <- 'OS'
    return(rbind(dt1, dt2))
  }))
}))
parallel::stopCluster(cl)

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.cohort_based_metrics.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),        
       plot = Fig, units = 'in', 
       width = 10, height = 10)
Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B). Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) based on the top 10 LFs/PCs derived from each tool. C) Same as in panels A and B), but split by the value of k used in clustering. D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival or Progression Free Survival endpoints."
```

@@

# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
event <- 'PFI'
time <- 'PFI.time'

@@

dt <- do.call(rbind, lapply(names(LFs), function(tool) {
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]

  # find LFs predictive of MSI 
  dt <- get_factor_specific_variables(M, y)
  dt$tool <- tool
  return(dt)
}))

# compare top predictive factors across tools
dt$tool <- toupper(gsub("_factors", "", dt$tool))
p1 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
       aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5))

# focus on maui LFs to visualize MSI status-related LFs
M <- LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']]
y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
M <- M[!is.na(y),]
y <- y[!is.na(y)]
top_vars <- unique(get_top_by_label(dt[tool=='MAUI'][padj < 0.05][order(padj)], 'ref_cl', 10)[['variable']])
df <- plot_tsne(M[,top_vars], y, returnData = T)
top_lf <- dt[order(padj)][ref_cl == 'MSI-H']$variable[1]
df <- merge(df, data.frame(M[,top_lf,drop =F]), by = 'row.names')
p2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
    geom_point(aes(color = as.factor(group))) + 
    theme(legend.title = element_blank())
p3 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
    geom_point(aes_string(color = top_lf)) + 
    scale_color_gradient(low = 'gray', high = 'red')

p4 <- ggboxplot(df, x = 'group', y = top_lf, color = 'group', 
          add = 'jitter') + theme(legend.position = 'none', 
                                  axis.title.x = element_blank())

# plot a roc curve
rocs <- predict_cluster_labels_glmnet(M, y, returnROC = TRUE)
p5 <- plot_roc_list(rocs)

@@

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p6 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + 
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45)) + 
  labs(x = paste0("top features contributing to ",top_lf))

Fig <- cowplot::plot_grid(p1, p5, p2, p3, p4, p6,
                          nrow = 3)
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.MSI.status.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 8, height = 8)
Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors. A) AUC values for predicting MSI status using MAUI LFs B) Top features that contribute to MAUI latent factor (‘LF95’) that is the most predictive latent factor in distinguishing samples with high microsatellite instability from the rest. C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor (‘LF95’) which is most predictive of MSI-H samples."
```
